### 2017-04-08 Manuscripts/cog-psych/cogpsych.tex
quick edits to the jk13 analysis section

Although existing accounts of creative generation broadly overlook the role of category contrast in determining what is novel versus familiar, effects of category contrast are undoubtedly ubiquitous within the published data. In each study, participants were \textit{successful} in creating something new: the question is about \textit{how} they were able to do so. To provide some concrete evidence of these effects within existing datasets, we conducted an analysis of the data from Experiment 3 of the report by \cite{jern2013probabilistic}.

\nbcnote{Report analyses on the \cite{jern2013probabilistic} data here...}

>>>>

Although existing accounts of creative generation broadly overlook the role of category contrast in determining what is novel versus familiar, these effects are undoubtedly ubiquitous within the published data. In each study, participants were \textit{successful} in creating something new: the question is about \textit{how} they were able to do so. To provide concrete evidence of  effects of category contrast within existing datasets, we conducted an analysis of Experiment 3 from \cite{jern2013probabilistic}.

Participants in their experiment were exposed to members of two experimenter-defined categories of `crystals' varying in hue, saturation, and size. Each category possessed a unique hue, but varied in saturation and size: In the `Positive' condition, there was a positive correlation between these features (i.e., larger sized crystals were more saturated), and in the `Negative' condition this relation was reversed. In the `Neutral' condition, there was no correlation between saturation and size. After learning about the categories from each condition, participants were asked to generate six exemplars belonging to a novel class. As noted above, \cite{jern2013probabilistic} found that the generated categories tended to follow the distributional properties of the experimenter-defined categories: Generated categories were tightly distributed along the hue feature, and possessed the same saturation-size correlations as in the learned categories. \cite{jern2013probabilistic}, however, did not provide an analysis of how the generated categories \textit{differed} from the experimenter-defined categories.

Because each experimenter-defined category in the \cite{jern2013probabilistic} experiment possessed a distinct hue, it is sensible to propose that participants would make their generated category distinct by using hues that differ from those present in the experimenter-defined category: If creative generation is influenced by category contrast, the hues of generated categories should be systematically different from those of the experimenter-defined categories. Problematically, however, stimulus hue was measured on an circular (HSV) scale with limits $[0,1]$: both 0 and 1 corresponded to a `red' hue, and all other possible hues were represented between these values. By consequence, proximity on the hue dimension does not directly correspond to perceptual similarity, and we cannot simply compare the hue values to determine the extent of the differences.

To overcome this issue, we categorized the hues of each participant's generated categories into one of six uniformly-spaced color groups: \{\textit{red}: 0 \& 1, \textit{yellow}: 0.167, \textit{green}: 0.333, \textit{teal}: 0.5, \textit{blue}: 0.667, \textit{pink}: 0.833\}. Each generated crystal was placed into the color group closest to its hue value. After each example's hue was categorized, we computed the amount of overlap between the hues of the six hues generated by each participant and the two hues present in the experimenter-defined categories: If contrast influences the hues of the generated categories, we should observe minimal overlap between the generated and experimenter-defined categories.

\nbcnote{Joe should probably do a pass through that language. Maybe add a cite for proximity-similarity distinction?}

\begin{figure}
    \begin{center}
        \inputpgf{figs/}{jk13-huecontrast.pgf}
        \caption{Analysis of data from \cite{jern2013probabilistic}, Experiment 3. Plotted is the number of generated items that share a color group with one of the experimenter-defined classes. The ``Expected'' data follows a Binomial distribution with $p=2/6$, given there were two experimenter-defined classes, and 6 color groups.}
        \label{fig:jk13-huecontrast}
    \end{center}
\end{figure}

These data, shown in Figure \ref{fig:jk13-huecontrast}, reveal a clear pattern: The majority of participants generated categories possessing entirely distinct hues; with 0/6 exemplars sharing a hue with the experimenter-defined categories. These results can be compared to the predictions of a Binomial model, which proposes that participants generate hues at random. According to the model, if hue selection is not systematic, the probability that any given example will lie in the same color group as an experimenter-defined category is $p = 2/6$, as there were two experimenter-defined categories and six possible color groups. Thus, the number of shared hues is Binomially distributed with $p=2/6$ and $n=6$ (as participants generated six examples in each condition). Comparison of the behavioral results to the model predictions reveals that participants were far more likely to select distinct hues: participants tended to generate items that were perceptually distinct from the categories they had learned.

These results are notable in that they provide some early, concrete evidence that creative generation is influenced by category contrast. Taken alongside the analyses reported by \cite{jern2013probabilistic}, our results suggest that generated categories tend to be distinct from \textit{and} distributionally similar to what is already known. However, while these results indicate that contrast exerts \textit{some} influence, they do not precisely describe the nature of that influence. In the sections below, we propose a quantitative framework specifying the role of category contrast in creative generation.
